The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-03-01 src/ms.tex
draft subsection on precision
Signed-off-by: nstarman <nstarkman@protonmail.com>

The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citealt{astropy:2018}.
% feels like ↑ could be expanded to 2 sentences

% Maybe briefly outline the sectors of  
% Using \astropypkg for
% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
% Astropy has grown to interoperate, e.g. NumPy

Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.

% Summarize major pieces of development since the last paper (since v2.0). A few
% ideas (taken from What's New pages) below, but feel free to expand on or remove
% from this list:
% \begin{itemize}
%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
%     \item APE 14 WCS
%     \item A new SpectralCoord class for representing and transforming spectral quantities
%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
% \end{itemize}

\subsubsection*{New Features} \label{sec:core-features-new}

\begin{itemize}
%  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
%  \item Uncertainties and Distributions
  \item ERFA module is now it’s own thing; Obliquely tie in that units
    remains in astropy
  \item ? Pytest plugin ecosystem
%  \item A new SpectralCoord class for representing and transforming spectral quantities
%    Velocity in Coordinates
  \item WCS API : A shared Python interface for World Coordinate Systems
        Include APE reference
\end{itemize}

New features come in various forms, from new submodules providing
wholly new capabilities, to larger additions to existing submodules,
to large rewrites of the interface of modules.  This range is spanned
by the four main features that entered the astropy core, which we
describe below.

@@

@@

>>>>

  The relationship between \astropy and community is reciprocal, with the
  growing and evolving needs of the community and the maturation of the broader
  scientific \python ecosystem now driving the development of the \astropypkg
  core package.

  One manifestation of this development realizing some features do not actually
  belong in the core package, either because they are of more general use
  outside of \astropypkg, or because they are too specialized. An example of the
  former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA)
  software Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that
  \astropy carried, as well as the \python wrappers for it.  Since this is more
  basic infrastructure, with a release cycle set by SOFA, it made more sense to
  create separate packages that \astropypkg can depend on: \package{ERFA}
  \citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
  was too specialized was a Virtual Observatory sub-package (now removed).  Of
  its content, the Simple Application Message Protocol made sense to keep
  (\astropysubpkg{samp}), but other parts belonged more properly in {\tt
  Astroquery} (see Sect.~\ref{sec:astroquery}).

  Within \astropypkg, development may be roughly split into a few categories:
  new features added, performance enhancements, intra and inter-package
  interoperability, and improvements to precision, accuracy, and
  reproducibility. We discuss these in turn.

\subsubsection*{New and Planned Features} \label{sec:core-features-new}

% \begin{itemize}
%  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
%  \item Uncertainties and Distributions
%   \item ERFA module is now it’s own thing; Obliquely tie in that units
%    remains in astropy -> above
%  \item ? Pytest plugin ecosystem -> done elsewhere
%  \item A new SpectralCoord class for representing and transforming spectral quantities
%    Velocity in Coordinates
%  \item WCS API : A shared Python interface for World Coordinate Systems
%        Include APE 14 reference
% \end{itemize}

New features come in various forms, from new submodules providing
wholly new capabilities, to larger additions to existing submodules,
to large rewrites of the interface of modules.  This range is spanned
by the five main new features that entered the astropy core since the
previous summary \citepalias{astropy:2018}.

New features present a particular difficulty for a mature package such
as \astropypkg, because by their nature they have not been tried out
much and it is difficult to know in advance what the best interface
will be, and hard to address all possible use cases in one go, even
within \astropypkg\ itself.  To give a sense of this, we include in
our descriptions how we hope the new features will evolve (realizing
that, in the end, in a user-driven project like \astropy, what happens
depends mostly on what is highest on the priority list of those that
have time to contribute).

@@

\paragraph{A High-Level Interface to World Coordinate Systems.}
Astronomical data are often provided with information about the
``real-world'' coordinates that correspond to pixel coordinates. This
mapping is the essence of the ``World Coordinate System (WCS)''
concept.  From its inception, the \astropysubpkg{wcs} sub-package
allowed access to WCS information provided in, e.g., FITS files, but
it became clear that other WCS standards and representations had to be
supported for new missions and observatories (e.g., the James Webb
Space Telescope and the Rubin Observatory).  To harmonize these, a new
high-level interface was created, based on a formal design first
proposed via an \astropy\ Proposal for Enhancement (\astropyAPE{14}).
The hope is that by having a formal design, it will be straightforward
for other packages implementing WCS objects to either modify their
classes to conform to the new interface or build thin wrappers that
conform to it.  The implementation in \astropysubpkg{wcs} interacts
well with other \astropypkg\ objects such as \astropySkyCoord\ and
\astropyTime.

@@

\subsection*{Precision, Accuracy, and Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}

  An important focus of \astropy has been to increase the precision and accuracy
  of its functionality, while, where appropriate, making sure results obtained
  using older versions of \astropypkg are reproducible.
  % Something about ScienceState
  Notable improvements have been made to \astropytime, \astropyconstants,
  \astropycoordinates, and \astropycosmology.

  \paragraph{Time}

    In astronomy, time accuracy down to the second is frequently important.
    Whether for planning observations or crunching data, missing seconds can
    meaningfully impact results. Before, when using \astropyTime in \astropytime
    leap seconds had to be manually applied; now leap seconds are applied
    automatically. Moreover, \astropypkg updates internal time data-files to
    ensure that the correct leap second adjustment is always used. For
    reproducibility, e.g. with an old code, \astropyLeapSeconds may also be
    manually applied.

  \paragraph{Constants}

    Measurements of physical constants (and the units defined from them) improve
    over time. Periodically the standardized systems of units and constants are
    updated to reflect these improvements, for instance in 2019 the SI system was
    redefined \citep{NIST2019}, with an accompanying update to the physical
    constants in
    \href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
    2018}}. \astropypkg contains two modules -- \astropyconstants and
    \astropyunits -- for working with constants and units, respectively.
    \astropyconstants now defaults to use the \texttt{SI/CODATA 2018} values, with
    the relation between units in \astropyunits based upon these constants. Most
    of \astropypkg and affiliate packages build upon \astropyunits, so this update
    has far-reaching effects across the \astropy ecosystem and astronomy
    community.

    For reproducibility, \astropypkg allows the constants (and therefore units)
    definitions to be rolled back to prior values, for instance the
    \texttt{SI/CODATA 2014} definitions.
    % CITATION
    The constants and units definitions are set by configuration file as well as
    \astropyScienceState (an runtime configuration controller). For work where
    updates to the values of the fundamental constants significantly alters
    results we recommend including with the work an astropy configuration file,
    which specifies the set of constants used.

  \paragraph{Cosmology}
    % TODO! At least one good sentence about "what is the benefit of astropy.cosmology"

    \astropycosmology contains classes for representing cosmological models.
    % currently all homogeneous and isotropic variations of the
    % Friedmann-Lemaitre-Robertson-Walker metric.
    Bundled with the classes are commonly used \astropyCosmology realizations, for
    example best-fit measurements from
    \href{https://lambda.gsfc.nasa.gov/product/map/current/}{WMAP}
    \citep{WMAP2003} and \href{https://www.nasa.gov/mission_pages/planck}{Planck}
    \citep{PlanckMission:2006}.

    When these important missions publish new measurements, \astropycosmology is
    updated to include these results as realizations of the appropriate class
    (generally \astropyFlatLambdaCDM). The models used by WMAP and Planck do not
    always exactly correspond to \astropycosmology classes, for instance the
    Planck 2018 results \citep{Planck2018VI} include massive neutrinos in
    $\Omega_{matter,0}$, while in \astropyFlatLambdaCDM this mass contribution is
    stored in a separate parameter. Consequently, while some parameter values
    appear different from the source paper, the \astropyCosmology realizations
    correctly reproduce the cosmological models.

    Since \citealt{astropy:2018} the following cosmology realizations have been
    added:
    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP1.html}{\texttt{WMAP1}}
    from WMAP First Year \citep[Table 7: WMAP + CBI + ACBAR + 2dFGRS +
    Lya]{WMAP1Year:2003};
    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP3.html}{\texttt{WMAP3}}
    from WMAP Three Year \citep[Table 6: WMAP + SNGold]{WMAP3Year:2007};
    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck15.html}{\texttt{Planck15}}
    from Planck 2015 \citep[Table 4: TT, TE, EE + lowP + lensing +
    ext]{Planck2015XIII:2016};
    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck18.html}{\texttt{Planck18}}
    from the Planck 2018 best-fit cosmological parameters \citep[Table 2: TT, TE,
    EE + lowE + lensing + BAO]{Planck2018VI:2020}.

    Astropy provides a
    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.default_cosmology.html}{configurable
    default cosmology}. All calculations done in a cosmological context will use
    this default cosmology. The default \astropyCosmology has been updated to the
    Planck 2018 parameters. For reproducibility, the default cosmology may be set
    to old defaults, such as the WMAP Three Year values. The default cosmology is
    dynamically configurable using \astropyScienceState, meaning a set of
    calculations may be run using different assumed cosmologies, and results
    compared between the two.
    % Repeat the word default less.

  \paragraph{Coordinates}

    Being able to describe the position of objects, i.e. their coordinates, is
    fundamental across many knowledge domains. \astropypkg provides
    \astropycoordinates for working with low-level positional and velocity data
    all the way to high-level coordinate objects with reference frames,
    atmospheric information, and more. A central feature of \astropycoordinates is
    the ability to transform data between reference frames, e.g. \astropyICRS
    \citep{ICRS:1997} to \astropyGalacticLSR \citep{GalacticLSR:2010}. However,
    there have been some historical limitations for spatially proximate
    transformations, impacting the usefulness of \astropycoordinates for
    ground-based telescopes and intra-solar astronomy. Transformations in the
    \astropyAltAz frame were reasonably precise for very distant objects, but
    wrong by up around several arcseconds for e.g. the location of the moon. Now
    these transformations are much more precise, down to the milli-arcsecond
    level. Similar precision improvements were made to the
    \astropyapi{coordinates.builtin_frames.HADec}{Hour Angle-Declination} frame
    transformations. Additionally,
    \astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
    and associated transformations have been updated to correctly reflect the
    “true” (includes precession but not nutation) and “mean” (precession and
    nutation) terminology.

    For Galactic astronomers, the \astropyGalactocentric
    \astropyapi{coordinates.galactocentric_frame_defaults}{frame defaults} have
    been updated to include more recent measurements (tabulated in the
    \href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html#astropy.coordinates.galactocentric_frame_defaults.references}{\texttt{frame\_defaults.references}}
    attribute). For reproducibility, old definitions are still available, and will
    continue to be so as \astropypkg is updated with forthcoming measurements of
    the Galactic center, solar motion, etc.
    % may be used by setting the default value to ``pre-v4.0'' in the
    % \href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html}{\texttt{galactocentric\_frame\_defaults}}
    % \astropyScienceState.
